### 2018-02-06 LCAforActiveSonarImaging/submission/submit2/answer/answer.tex
Major commit before messing with PhD thesis

\begin{enumerate}
\item The steering bound $\phi_\mathrm{3dB}$ depends on not just $\beta$, but also on the number of elements $M$ and the element spacing relative to the wavelength $\frac{d}{\lambda}$. 
\begin{enumerate}
\item We have supplied additional info on how the steering is affected by these parameters in appendix C.
\item Equations (12) and (14) only gave the correct results for our system configuration, and were not very intuitive. Appendix C now replace and extends these equations, and we removed them in hope this will make the article easier to read.
\end{enumerate}
\item Various spelling flaws attempted corrected.
\item Appendix C added. Contains information of where source code and media files can be located. Before the article goes to publishing we will make sure that this link is permanent.
\item Reference to Synnevåg's proceedings article from 2008, ``A low-complexity data-dependent beamformer'' replaced with his 2011 journal article with the same name. We never intended to reference the proceedings article since this work extends his journal version.
\end{enumerate}

@@

\begin{enumerate}
% \emph{General}
\hi{Abstract}
\qi{"the array's spatial covariance matrix" might be a bit shocking for the non-expert Reader $\rightarrow$ The formulation in the Introduction might fit better "the influence of noise ..."}
\ai{This is a good point. We have reformulated the sentence to speak of computing the spatial statistics instead.}
%
\hi{Page 2}
\qi{"6 of which were micro-steered" $\rightarrow$ a short explanation or reference might help to understand what is meant by "micro-steered"}
\ai{Another good point. This wording is now changed and a sentence elaborating on its meaning is added.}
%
\qi{"it was not clear how how many" $\rightarrow$ delete 1 x "how"}
\ai{Thank you. Corrected.}
%
\qi{"with an element beamwidth" $\rightarrow$ what is meant by element beamwidth? Maybe an additional figure could be used to explain this.}
\ai{Good remark. We meant the -3dB opening angle of the elements. We have reworded this the two places it was mentioned.}
%
\hi{Page 3, after (1)}
\qi{"Applying a window that trails off .... Minimum variance distortion-less response (MVDR) beamformer." $\rightarrow$ real and complex Windows? The introduction should be more formal: x ist introduced as delayed data, weights or delays to change, or both is explained later. $\rightarrow$ The difference between noise and sidelobes should be pointed out.}
\ai{Good observations. We realize this was ambiguous and have now rewritten this paragraph in hopes it will be clearerr.}
%
\hi{Page 5, equation (5)}
\qi{The lateral correlation and the reason why equation (5) is a good choice should be better explained.}
\ai{Indeed. We have rewritten the paragraph slightly, but the technique is mentioned mostly for completeness. It wasn't used.}
%
\hi{Page 6, last row}
\qi{"they are optimal for any value" $\rightarrow$ optimal in which sense?}
\ai{Thank you. This was mentioned earlier in the same paragraph, just after (8). Is this insufficient?}
%
\hi{Fig 3}
\qi{why is a 40x40 pixel group a good choice? could the associated broadside direction be made visible in the figure, too?}
\ai{This is an excellent question. This should perhaps be related to the detail level in the image, which in turn depends on the array's imaging performance. We took an ad-hoc approach to this and selected 40x40 pixels to get sufficient statistics for a visually good looking histogram. We have now written this into the caption of Fig. 3.}
%
\hi{Page 8}
\qi{reference to Fig.1 $\rightarrow$ could the sequence of figures be changes to have Fig. 1 closer to this paragraph}
\ai{We agree this would be better. In our two column print the figure is on the page before its first reference. We assume this will be placed properly when processed for publishing.}
%
\hi{Page 11}
\qi{where is the synthetic aperture needed/applied in this work, or is it just the name of the array? how does the work on LCA/MVDR Change the Quality of synthetic aperture? What exactly is the sectorscan mode?}
\ai{The reviewer is quite right, it is a system designed for SAS imaging, but we only create image sectors for its data. That is, the images are from single ping, rather than synthesized from several pings. We agree this was worded in an ambiguous way and have now made relevant changes, as well as added a remark of what it meant by sectorscan.}
%
\end{enumerate}

>>>>

\begin{enumerate}
\item The steering bound $\phi_\mathrm{3dB}$ depends on not just $\beta$, but also on the number of elements $M$ and the element spacing relative to the wavelength $\frac{d}{\lambda}$. 
\begin{enumerate}
\item We have supplied additional info on how the steering is affected by these parameters in appendix A.
\item Equations (12) and (14) only gave the correct results for our system configuration, and were not very intuitive. Appendix C now replace and extends these equations, so removed them in hope this will make the article easier to read.
\end{enumerate}
\item Various spelling flaws corrected.
\item Appendix C added. Contains information of where source code and media files can be located. Before the article goes to publishing we will make sure that this link is permanent.
\item Reference to Synnevåg's proceedings article from 2008, ``A low-complexity data-dependent beamformer'' is replaced with his 2011 journal article with the same name. This was an error from our side, we never intended to reference the proceedings article since this work extends his journal version.
\end{enumerate}

@@

\begin{enumerate}
% \emph{General}
\hi{Abstract}
\qi{"the array's spatial covariance matrix" might be a bit shocking for the non-expert Reader $\rightarrow$ The formulation in the Introduction might fit better "the influence of noise ..."}
\ai{This is a good point. We have reformulated the sentence to speak of computing the spatial statistics instead.}
%
\hi{Page 2}
\qi{"6 of which were micro-steered" $\rightarrow$ a short explanation or reference might help to understand what is meant by "micro-steered"}
\ai{Another good point. This wording is now changed and a sentence elaborating on its meaning is added.}
%
\qi{"it was not clear how how many" $\rightarrow$ delete 1 x "how"}
\ai{Thank you. Corrected.}
%
\qi{"with an element beamwidth" $\rightarrow$ what is meant by element beamwidth? Maybe an additional figure could be used to explain this.}
\ai{Good remark. We meant the -3dB opening angle of the elements. We have reworded this the two places it was mentioned.}
%
\hi{Page 3, after (1)}
\qi{"Applying a window that trails off .... Minimum variance distortion-less response (MVDR) beamformer." $\rightarrow$ real and complex Windows? The introduction should be more formal: x ist introduced as delayed data, weights or delays to change, or both is explained later. $\rightarrow$ The difference between noise and sidelobes should be pointed out.}
\ai{Good observations. We realize this was ambiguous and have now rewritten this paragraph in hopes it will be clearerr.}
%
\hi{Page 5, equation (5)}
\qi{The lateral correlation and the reason why equation (5) is a good choice should be better explained.}
\ai{Indeed. We have rewritten the paragraph slightly, but the technique is mentioned mostly for completeness. It wasn't used.}
%
\hi{Page 6, last row}
\qi{"they are optimal for any value" $\rightarrow$ optimal in which sense?}
\ai{Thank you. This was mentioned earlier in the same paragraph, just after (8). Is this insufficient?}
%
\hi{Fig 3}
\qi{why is a 40x40 pixel group a good choice? could the associated broadside direction be made visible in the figure, too?}
\ai{This is an excellent question. This should perhaps be related to the detail level in the image, which in turn depends on the array's imaging performance. We took an ad-hoc approach to this and selected 40x40 pixels to get sufficient statistics for a visually good looking histogram. We have now written this into the caption of Fig. 3.}
%
\hi{Page 8}
\qi{reference to Fig.1 $\rightarrow$ could the sequence of figures be changes to have Fig. 1 closer to this paragraph}
\ai{We agree this would be better. In our two column print the figure is on the page before its first reference. We assume this will be placed properly when processed for publishing.}
%
\hi{Page 11}
\qi{where is the synthetic aperture needed/applied in this work, or is it just the name of the array? how does the work on LCA/MVDR Change the Quality of synthetic aperture? What exactly is the sectorscan mode?}
\ai{The reviewer is quite right, it is a system designed for SAS imaging, but we only create image sectors for its data. That is, the images are from single ping, rather than synthesized from several pings. We agree this was worded in an ambiguous way and have rewritten it for improved clarify, as well as elaborated on the meaning of sectorscan imaging.}
%
\end{enumerate}

### 2018-02-06 Simulator/article.bak.tex
Major commit before messing with PhD thesis

@@

% - Adapt simulation to its behaviour, see if they still match.

\IEEEPARstart{A}{utonomous} underwater vehicles (AUVs) are becoming increasingly important in human underwater endeavors. When equipped with a synthetic aperture sonar (SAS), they can e.g. be used to map the seafloor, survey underwater biology, help localize wrecks, inspect oil pipes, fields and electricity lines, or hunt for mines and biohazard waste. Such independent systems have much to gain from being able to adapt to the scenario at hand. For instance, with a well designed automatic target recognition (ATR) system on board they can adapt their mission plan to collect more information on particularly interesting objects, or fine-tune the sonar and navigation system to improve imaging performance.

% Automatic target recognition (ATR) is an important component in autonomous underwater vehicles (AUVs), as it allows the vehicle to adapt its mission plan to e.g. revisit detected objects for closer examination. One way to classify these objects is by using template matching. The principle involved  is to isolate an image segment containing the object of interest, compare it with a set of template images of the relevant object classes, and assign it to the class of the template with the the best fit.

In conventional ATR an object of interest is first isolated, its properties attempted estimated, and a classifier used to determine what object class these properties belong to. The challenge is finding unique and intrinsic object properties, such as its type, shape and size, and to do so independently of extrinsic properties such as its alignment, burial depth, degeneration level or view angle. While SAS provide images with state of the art resolution and dynamic range, they still only represent an ambiguous acoustical 2D projection of the 3D object, littered with a highly coherent noise floor.

An alternative approach is to use template matching, where the isolated image segment is compared to a set of template images covering the relevant object classes, and assigned to the class of the template with the best fit. Since the templates are computed from 3D models, this reduces the issue of image ambiguity. However, a challenge is finding a set of templates that represent the actual configurations of object and seafloor, Another issue is that the size of precomputed library of templates grows exponentially with the number of parameters needed. This limits its use to only a few parameters being coarsely sampled, ultimately yielding inaccurate results~\cite{Midelfart2010}.

% We avoid the problems of conventional template matching by computing the templates on-the-fly as part of the classification process, thereby not needing a precomputed set of templates at all. 
%  that avoids this problem by running sufficiently fast to create templates adapted to the actual scene as part of the classification process.
% Classification process

We tackle the problems of conventional template matching by performing:
%
\begin{itemize}
\item \emph{Adaptive template matching}; where a template is selected based on object and seafloor parameters estimated from the SAS image~\cite{Midelfart2010}, and
\item \emph{Real-time template simulation}; where a template is created tailored to the estimated parameters, improving its accuracy and avoiding the need of a precomputed template library.
\end{itemize}
%

The methods presented herein will focus on the template simulator, but some details will also be given on adaptive template matching for completeness. More details on this technique can be found in~\cite{Midelfart2010}.

 This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering. The simulator loads a 3D model of the seafloor and object class into OpenGL and positions the camera at the sonar location. The scene is highlighted using a light tube that extends to infinity along the AUV's axis of probagation. We assume rough, isotropic surfaces reflecting sound energy equally in all directions. This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface. The OpenGL rendering produces an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis. For maximum flexibility this data is finally combined with OpenCL to produce the image templates.

as part of the classification process. This is achieved with the aid of the massive computing power in graphics processing units (GPUs) and optimized software libraries for scene rendering.

The simulator loads a 3D model of the seafloor and an object class model into OpenGL, where emitted sound waves are modeled with a cylindrical light source placed at the sonar transmit location.

We assume rough, isotropic surfaces reflecting sound energy equally in all directions. This can be modeled with a Lambertian scattering model~\cite{Blake1993,Bell1995}, where the intensity of the backscattered sound solely depends on the incidence angle of the transmitted signal onto the model surface. When rendering OpenGL is set up to produce an optical 2D intensity image and depth map that reveals the distance from each pixel to the propagation axis. For maximum flexibility this data is finally combined with OpenCL to produce the image templates.

Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}. However, most implementations have prioritized accurate acoustic modeling at the expense of execution speed. One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b}. Similarly to our approach it takes advantage of parallel processing on GPUs, and it uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering. However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation. Also, their simulator superimpose images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with greater flexibility.

For lack of a better name, the simulator will be referred to as "FFISim" throughout this article.

Show:
\begin{itemize}
\item Adaptive template matching to avoid large static template libraries,
\item Improved classification performance by feeding simulator with parameters derived off SAS image,
\item Further improved classification by performing a fine-search within a narrow search space around the SAS derived parameters. GPU makes this more powerful.
\end{itemize}

\section{Methods}

Creating a 2D view of an arbitrary complex 3D scene is a non-trivial matter.  This is why we decided to use the core OpenGL pipeline to do this for us. OpenGL is a popular, well-matured and multi-platform application programming interface (API) for rendering 2D and 3D vector graphics. It relieves us from the intricacies of projecting vertices, faces and textures defined in a 3D space onto a suitable 2D image plane. This section explains how we set up OpenGL for this task, and then proceeds to describe how we post process the OpenGL images with OpenCL to form the final sonar image.

\begin{itemize}
\item Load arbitrary complex models
\item Simulate sonar with OpenGL
\item Speed considerations
\end{itemize}

Models are not ROCCAN (???)

\subsection{Loading the scene}

Our models are stored in regular 3D model files. The simulator begins by loading these into a tree of nodes, each representing a vertex in the model with corresponding facet and texture information. Then the vertices are placed into a virtual scene in OpenGL, and a camera set up to capture the part of the scene corresponding to the image. When OpenGL renders the image, it first computes which vertex contributes to which pixel in the image, then computes the intensity and color of the pixel based on the node's facet, texture and scattering model. The code that operates on vertices are called a vertex shader, and the code that operates on pixels are called a fragment shader.

The mapping of the vertices from object coordinates to image coordinates happen through a set of transformations:
%
\begin{align*}
\bmat{x\\y\\0\\1}_\text{image} = \boldsymbol{P} \cdot \boldsymbol{V} \cdot \boldsymbol{M} \cdot \bmat{x\\y\\z\\1}_\text{object},
\end{align*}
%
where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.

The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
%
\begin{align}
\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
\end{align}
%
where $\boldsymbol{S}$, $\boldsymbol{R}$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively. These are provided for reference in Appendix \ref{transformation_matrices}.

@@

\section{Transformation matrices}\label{transformation_matrices} 
%
We place, resize and orient objects in an euclidean 3D space using suitable transformation matrices that are applied to each of the object's vertices:
\begin{align}
\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
\end{align}
Here a model vertex at position $(x,y,z)$ is first scaled with $\boldsymbol{S}$, then rotated with $\R$
There is nothing special to these matrices, but we supply them here for completeness.

>>>>

  % \newcommand\includegraphics[2][]{%
  % \immediate\write18{./bin/laFigure #2 #1}%
  % \input{result}}%

@@

% - Adapt simula th\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
\end{align}
%
where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively. These are provided for reference in Appendix \ref{transformation_matrices}.

@@

\section{Transformation matrices}\label{transformation_matrices} 
We place, resize and orient objects in an euclidean 3D space using suitable transformation matrices that are applied to each of the object's vertices:
\begin{align}
\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
\end{align}
Here a model vertex at position $(x,y,z)$ is first scaled with $\boldsymbol{S}$, then rotated with $\R$
There is nothing special to these matrices, but we supply them here for completeness.

### 2018-02-06 Simulator/article.tex
Major commit before messing with PhD thesis

The mapping of the vertices from object coordinates to image coordinates happen through a set of transformations:
%
\begin{align*}
\bmat{x\\y\\0\\1}_\text{image} = \boldsymbol{P} \cdot \boldsymbol{V} \cdot \boldsymbol{M} \cdot \bmat{x\\y\\z\\1}_\text{object},
\end{align*}
%
where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.

The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
%
\begin{align}
\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
\end{align}
%
where $\boldsymbol{S}$, $\boldsymbol{R}$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively. These are provided for reference in Appendix \ref{transformation_matrices}.

Next we map the model into world coordinates using the view matrix $\V$. This involves aligning the camera to precisely fit the image region. In our case the camera represents the phase center of the sonar, and it is placed in the world origin. The transformation then take the form of
%
\begin{align}
\V &= \mat{T}(x_\text{c}, y_\text{c}, z_\text{c}) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
&= 
\bmat{
1  &  0  &  0  &  x_\text{c} \\
0  &  1  &  0  &  y_\text{c} \\
0  &  0  &  1  &  z_\text{c} \\
0  &  0  &  0  &  1 \\
} \cdot
\bmat{
1  &  0           &  0           &  0 \\
0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
0  &  0           &  0           &  1 \\
},
\end{align}
%
where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\phi_x$ is the elevation angle\todo{is this correct?} of the sonar. It is given as
%
\begin{align*}
\phi_x &= \arctan\left(\frac{y_\text{mean}}{-z_\text{mean}}\right).
\end{align*}\todo{use absolute sign?}

@@

\section{Transformation matrices}\label{transformation_matrices} 
%
We place, resize and orient objects in an euclidean 3D space using suitable transformation matrices that are applied to each of the object's vertices:
\begin{align}
\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T}\R\boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model}.
\end{align}
Here a model vertex at position $(x,y,z)$ is first scaled with $\boldsymbol{S}$, then rotated with $\R$
There is nothing special to these matrices, but we supply them here for completeness.

We implement rotations in a 3D euclidean space by combining matrices that rotate around each of the three respective dimensions. For rotation by an angle $\phi_x$ clockwise around the $x$-axis we have
\begin{align}
\R_x(\phi_x) &= \bmat{
1  &  0           &  0           &  0 \\
0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
0  &  0           &  0           &  1 \\
},
\end{align}
and similarly, for the other two euclidean dimensions we have
\begin{align}
\R_y(\phi_y) &= \bmat{
\cos\phi_y  &  0  &  \sin\phi_y  &  0 \\
0           &  1  &  0           &  0 \\
-\sin\phi_y &  0  &  \cos\phi_y  &  0 \\
0           &  0  &  0           &  1 \\
},
\end{align}
and
\begin{align}
\R_z(\phi_z) &= \bmat{
\cos\phi_z  &  -\sin\phi_z &  0  &  0 \\
\sin\phi_z  &  \cos\phi_z  &  0  &  0 \\
0           &  0           &  1  &  0 \\
0           &  0           &  0  &  1 \\
}.
\end{align}
The order in which these are applied matters. We rotate around the $x$-axis first, then around $y$ and finally around $z$:
\begin{align}
\R(\phi_x, \phi_y, \phi_z) &= \R_z(\phi_z)\;\R_y(\phi_y)\;\R_x(\phi_x).
\end{align}

\begin{align}
\boldsymbol{T}(dx,dy,dz) &= \bmat{
1  &  0  &  0  &  dx \\
0  &  1  &  0  &  dy \\
0  &  0  &  1  &  dz \\
0  &  0  &  0  &  1 \\
}
\end{align}

\begin{align}
\boldsymbol{S}(s_x,s_y,s_z) &= \bmat{
s_x  &  0    &  0    &  0 \\
0    &  s_y  &  0    &  0 \\
0    &  0    &  s_z  &  0 \\
0    &  0    &  0    &  1 \\
}
\end{align}

@@

% use section* for acknowledgement
\ifCLASSOPTIONcompsoc% % This command fixes abstract positioning for compsoc articles:
% \IEEEdisplaynotcompsoctitleabstractindextext
% 
% % (Optional) Add some extra info on cover page of peer review papers:
% % \ifCLASSOPTIONpeerreview
% % \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% % \fi
% 
% % Insert page break and insert second title (peer review mode)
% \IEEEpeerreviewmaketitle
% 
% 
%

  \section*{Acknowledgments}
\else
  \section*{Acknowledgment}
\fi

The authors would like to thank the Norwegian Defence Research Establishment (FFI) for funding to develop the simulator\todo{more?}.

@@

\input{bio/jo_inge_buskenes}
% \input{bio/jon_petter_aasen}
% \input{bio/carl-inge_nilsen}
% \input{bio/andreas_austeng}

@@

>>>>

Mapping of the vertices from object coordinates to image coordinates happen through a set of transformations:
%
\begin{align*}
\bmat{x\\y\\0\\1}_\text{image} = \boldsymbol{P} \cdot \boldsymbol{V} \cdot \boldsymbol{M} \cdot \bmat{x\\y\\z\\1}_\text{model},
\end{align*}
%
where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are model, view and transformation matrices, respectively. These will be described next.

The model matrix $\M$ is the first transform applied to the object, and is defined as:
%
\begin{align}
\M = \bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \cdot \bmat{x \\ y \\ z \\ w}_\text{model},
\end{align}
%
where $\boldsymbol{S}$, $\boldsymbol{R}$ and $\boldsymbol{T}$ are transformation matrices\todo{be explicit on dependencies?} that sets the object's scale, rotation and translation, respectively. Combined these transforms map the object from its local coordinate system to the world's coordinate system. They are provided for reference in Appendix \ref{transformation_matrices}.

Next the object is mapped into world coordinates with the view matrix $\V$. This involves aligning the camera to precisely fit the image region. In our case the camera represents the phase center of the sonar, and it is placed in the world origin. The transformation then take the form of
%
\begin{align}
\V &= \mat{T}(x_\text{c}, y_\text{c}, z_\text{c}) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
&= 
\bmat{
1  &  0  &  0  &  x_\text{c} \\
0  &  1  &  0  &  y_\text{c} \\
0  &  0  &  1  &  z_\text{c} \\
0  &  0  &  0  &  1 \\
} \cdot
\bmat{
1  &  0           &  0           &  0 \\
0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
0  &  0           &  0           &  1 \\
},
\end{align}
%
where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\phi_x$ is the elevation angle\todo{is this correct?} of the sonar. It is given as
%
\begin{align*}
\phi_x &= \arctan\left(\frac{y_\text{mean}}{-z_\text{mean}}\right).
\end{align*}\todo{use absolute sign?}

@@

\section{Transformation matrices}\label{transformation_matrices}

Each data model that we load into OpenGL is defined in its own local coordinate system. To map them into OpenGL world coordinates a set of common transformation matrices are applied to each model's vertices:
%
\begin{align}
\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \boldsymbol{R} \cdot \boldsymbol{S} \cdot \bmat{x \\ y \\ z \\ w}_\text{model}.
\end{align}
%
Here a model vertex at position $(x,y,z)$ is first scaled with $\boldsymbol{S}$, then rotated with $\boldsymbol{R}$ and finally translated with $\boldsymbol{T}$, resulting in the transformed coordinates $(x_t,y_t,z_t)$. The parameter $w$ defines whether the vertex is a position ($w=1$) or a direction ($w=0$), which need to be handled differently. While these matrices are well-known in the field of computer graphics, there exist several different conventions but we supply them here for completeness.

The scaling and translation matrices are simple:
%
\begin{align}
\boldsymbol{S}(s_x,s_y,s_z) &= \bmat{
   s_x  &  0    &  0    &  0 \\
   0    &  s_y  &  0    &  0 \\
   0    &  0    &  s_z  &  0 \\
   0    &  0    &  0    &  1 \\
  },
  \end{align}
%
and

TZF64LJO2AUK2XEM
%
\begin{align}
\boldsymbol{T}(dx,dy,dz) &= \bmat{
	1  &  0  &  0  &  dx \\
	0  &  1  &  0  &  dy \\
	0  &  0  &  1  &  dz \\
	0  &  0  &  0  &  1 \\
},
\end{align}
%
where ($s_x$, $s_y$, $s_z$) and ($dx$, $dy$, $dz$) is scaling and translation factors.

Rotations will be specified using intrinsic Euler angles, i.e. as yaw ($\phi$), pitch ($\theta$) and roll ($\psi$), representing clockwise rotations around the $z$-, $y$- and $x$-axis, respectively. The Brian-Tait $Z$-$Y'$-$X''$ convention is used for rotation order, in which roll is applied first, then pitch and finally yaw:
%
\begin{align}
\R(\phi_x) = \R_z(\phi)\R_y(\theta)\R_x(\psi),
\end{align}
%
where
%
\begin{align}
\R_x(\psi) &= \bmat{
1  &  0           &  0           &  0 \\
0  &  \cos\psi  &  -\sin\psi &  0 \\
0  &  \sin\psi  &  \cos\psi  &  0 \\
0  &  0           &  0           &  1 \\
},
\end{align}
%
\begin{align}
\R_y(\theta) & = \bmat{
\cos\theta	& 0          & \sin\theta & 0 \\
0					& 1                   & 0          & 0          &  \\
-\sin\theta  & 0                   & \cos\theta & 0          &  \\
0            & 0                   & 0          & 1          &  \\
},           &
\end{align}
%
and
%
\begin{align}
\R_z(\phi) &= \bmat{
\cos\phi  &  -\sin\phi &  0  &  0 \\
\sin\phi  &  \cos\phi  &  0  &  0 \\
0				&  0           &  1  &  0 \\
0           &  0           &  0  &  1 \\
}.
\end{align}
%
Note that at $\theta=\frac{\pi}{2}$ we have $\phi=-\psi$, and at $\theta=-\frac{\pi}{2}$ we have $\phi=\psi$. Hence, one degree of freedom is lost at these points, an effect known as Gimbal lock. This can be solved by avoiding Eulean angles, e.g. by direct specification of orientation in terms of a rotation axis and angle. This can be achieved using the Euler-Rodrigues formula or with quaternions. The latter is the most common, due to added benefits such as reduced computational complexity, improved numerical stability and simple ways to perform spherical interpolations. However, in for our application this is not needed.
%
Translation is performed last. 
%

@@

% use section* for acknowledgement
\ifCLASSOPTIONcompsoc% % This command fixes abstract positioning for compsoc articles:
% \IEEEdisplaynotcompsoctitleabstractindextext
% 
% % (Optional) Add some extra info on cover page of peer review papers:
% % \ifCLASSOPTIONpeerreview
% % \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% % \fi
% 
% % Insert page break and insert second title (peer review mode)
% \IEEEpeerreviewmaketitle
% 
% 
% 
  \section*{Acknowledgments}
\else
  \section*{Acknowledgment}
\fi

The authors would like to express their gratitude to the Norwegian Defence Research Establishment (FFI) for funding the development of the simulator.

@@

\input{bio/jo_inge_buskenes_new}
% \input{bio/jon_petter_aasen}
% \input{bio/carl-inge_nilsen}
% \input{bio/andreas_austeng}

@@

\input{derivations}

### 2018-02-06 Simulator/bio/jo_inge_buskenes_new.tex
Major commit before messing with PhD thesis

His research interests include radar and sonar technology, aptive image reconstruction, high performance computing, intelligent detector design and open source software.
\end{IEEEbiography}

>>>>

His research interests include radar and sonar technology, adaptive image reconstruction, high performance computing, intelligent detector design and open source software.
\end{IEEEbiography}

### 2018-02-06 Simulator/bio/roy_edgar_hansen_new.tex
Major commit before messing with PhD thesis

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bio/roy_color.eps}}]{Roy Edgar Hansen} (M'07) received the M.Sc. and Ph.D. degrees in physics from the University of Troms\o, Troms\o, Norway, in 1992 and 1999, respectively.

From 1992 to 2000, he was with the Norwegian research company TRIAD, working on multistatic sonar, multistatic radar, synthetic aperture radar (SAR), and underwater communications. Since 2000, he has been with the Norwegian Defence Research Establishment (FFI), Kjeller, Norway. He is currently Principal Scientist and team leader for the autonomous underwater vehicle development and the synthetic aperture sonar development at FFI. He is also Adjunct Associated Professor at the University of Oslo, Oslo, Norway.
\end{IEEEbiography}

>>>>

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{bio/roy_color.eps}}]{Roy Edgar Hansen} (M'07) received the M.Sc. and Ph.D. degrees in physics from the University of Troms\o, Troms\o, Norway, in 1992 and 1999, respectively.
The PhD thesis is named Measurements in the Mixed Layer by a Bistatic Multi-CW Doppler Sonar. 
From 1992 to 2000, he was with the Norwegian research company TRIAD, working on multistatic sonar, multistatic radar, synthetic aperture radar (SAR), and underwater communications. Since 2000, he has been with the Norwegian Defence Research Establishment (FFI), Kjeller, Norway. 
In the period 2009-2015, he was research manager for the autonomous underwater vehicle development at FFI, including the fields underwater navigation, battery and fuel-cell technologies, decisional autonomy, and synthetic aperture sonar. 
He is currently principal scientist at FFI. 
He is also adjunct associated professor at Department of informatics at University in Oslo, Norway.
His research interests include synthetic aperture sonar and radar, ultrasound imaging, and array signal processing.
\end{IEEEbiography}

### 2018-02-06 ThesisJo/introduction.tex
Major commit before messing with PhD thesis

\IEEEPARstart{O}{ne} of the greatest visions in science is that of artificial intelligence. If we could make machines not only do what we say, but also able to adapt their behaviour to handle the unexpected, the possibilities are perhaps only limited by our imagination.

>>>>

One of the greatest visions in science is that of artificial intelligence. If we could make machines not only do what we say, but also able to adapt their behavior to handle the unexpected, the possibilities are perhaps only limited by our imagination.
